---
title: "Nested Random Effects in lme4"
author: "Andrew L Jackson"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This example illustrates how the syntax for nesting random effects in `lme4::lmer` works. Essentially its a factor labelling behaviour.

```{r load pacakge}

# load the glmm package
library(lme4)

```

The point about data encoding in this example is that observations are nested within casks, which are nested within batches. However, the casks have been labelled as "a", "b" , "c" in each batch, even though there is no relationship between a cask labelled "a" in batch A with one labelled "a" in batch B. Hence the need to include the nesting structure via the "batch/cask" notation which is the same as including a batch:cask interaction, which is the same as labelling the casks uniquely as is done in the column "sample".

```{r data}
data("Pastes")

# view the data

Pastes

```

## Fit the random effects models

```{r fit}
# nested random effects as per the / notation
m1 <- lmer(strength ~ 1 + (1|batch/cask), data = Pastes)

# the same by manually specifying the interation term
m2 <- lmer(strength ~ 1 + (1|batch) + (1|batch:cask), data = Pastes)

# the same if one had encoded the data correctly in the first place...
m3 <- lmer(strength ~ 1 + (1|batch) + (1|sample), data = Pastes)
```

All three of these models are identical:

```{r output}
summary(m1)
summary(m2)
summary(m3)

```


